Analyzing the Red Wine Data Set by Jon Baker
========================================================

```{r echo=FALSE, message=FALSE, warning=FALSE, packages}

#Setting up libraries

library(ggplot2)
library(knitr)
library(dplyr)
library(gridExtra)
library(GGally)



```

```{r echo=FALSE, Load_the_Data}
# Load the Data

setwd('C:/Users/Home/Desktop/Project 4')
redwine <- read.csv('wineQualityReds.csv')

```

### Introducing the Red Wine Data Set

This data set looks at over 1,000 red wines. It looks like there are 13
variables to look at, with one of them being a primary key. I notice
that some variables operate on different orders of magnitude than others.

# Univariate Plots Section

Here are some summaries of the data set's variables.

```{r DataSummaries}

#These are the variables we are working with in the data set
names(redwine)

#Let's look at some examples of the different variables in action
str(redwine)

#Let's look at a summary of the different variables!

summary(redwine)

```
There is a primary key, 11 variables, and 1 quality variable that can be used
as a categorical variable.

Here are some charts looking at all of the variables in the data set.

```{r echo=FALSE, message=FALSE, warning=FALSE, Univariate_Plots}

p1 <- qplot(x = fixed.acidity, data = redwine,
            xlab = 'Fixed Acidity of Red Wines',
            ylab = 'Number of Red Wines',
            color = I('black'), fill = I('#990099'))
p2 <- qplot(x = volatile.acidity, data = redwine,
            xlab = 'Volatile Acidity of Red Wines',
            ylab = 'Number of Red Wines',
            color = I('black'), fill = I('#990099'))
p3 <- qplot(x = citric.acid, data = redwine,
            xlab = 'Amount of Citric Acid in Red Wines',
            ylab = 'Number of Red Wines',
            color = I('black'), fill = I('#990099'))
p4 <- qplot(x = residual.sugar, data = redwine,
            xlab = 'Residual Sugar in Red Wines',
            ylab = 'Number of Red Wines',
            color = I('black'), fill = I('#990099'))
p5 <- qplot(x = chlorides, data = redwine,
            xlab = 'Chlorides in Red Wines',
            ylab = 'Number of Red Wines',
            color = I('black'), fill = I('#990099'))

p6 <- qplot(x = free.sulfur.dioxide, data = redwine,
            xlab = 'Free Sulfur Dioxide in Red Wines',
            ylab = 'Number of Red Wines',
            color = I('black'), fill = I('#990099'))

p7 <- qplot(x = total.sulfur.dioxide, data = redwine,
            xlab = 'Total Sulfur Dioxide in Red Wines',
            ylab = 'Number of Red Wines',
            color = I('black'), fill = I('#990099'))

p8 <- qplot(x = density, data = redwine,
            xlab = 'Density of Red Wines',
            ylab = 'Number of Red Wines',
            color = I('black'), fill = I('#990099'))

p9 <- qplot(x = pH, data = redwine,
            xlab = 'pH of Red Wines',
            ylab = 'Number of Red Wines',
            color = I('black'), fill = I('#990099'))
p10 <- qplot(x = sulphates, data = redwine,
             xlab = 'Sulphates in Red Wines',
             ylab = 'Number of Red Wines',
             color = I('black'), fill = I('#990099'))
p11 <- qplot(x = alcohol, data = redwine,
             xlab = 'Alcohol in Red Wines',
             ylab = 'Number of Red Wines',
             color = I('black'), fill = I('#990099'))
p12 <- qplot(x = quality, data = redwine,
             xlab = 'Quality of Red Wines',
             ylab = 'Number of Red Wines',
             color = I('black'), fill = I('#990099'))


grid.arrange(p1,p2,p3,p4,p5,p6,p7,p8,p9,p10,p11,p12, ncol=5)

```

These visualizations were created in order to get a feel for the variables.
We can see that the "Quality of Red Wines" chart is based on integers, not real
numbers. "Fixed Acidity," "Density," "Volatile Acidity" and "pH" have the most
normal-looking curves out of the bunch. Many of these charts have tails on the
right, including the "Sulfur Dioxide" charts, "Sulphates" chart, "Alcohol" chart
and especially the "Residual Sugar" and "Chlorides" charts.

```{r echo=FALSE, message=FALSE, warning=FALSE, Univariate_Plots2}

ggplot(redwine, aes(x = alcohol)) +
geom_histogram(bindwith = .25, color = 'black') +
geom_vline(aes(xintercept = mean(alcohol)), color = 'red', linetype = 'dashed',
           size = 1)


```

Here is a close-up view of alcohol content. This is an important feature in
wines. We can see the mean is around 10.5 % or so. It will be interesting
to look this over in the multivariate analysis, as it is correlated with
quality.

```{r echo=FALSE, message=FALSE, warning=FALSE, Univariate_Plots3}

ggplot(redwine, aes(x = sulphates)) +
geom_histogram(bindwith = .25, color = 'black') +
geom_vline(aes(xintercept = mean(sulphates)), color = 'red', linetype = 'dashed',
           size = 1)


```

This is a histogram of sulfate content in red wines. You can see here that the
mean is around .7 mg/l or so. This will also be interesting in the multivariate
analysis, as it is correlated with quality.

```{r echo=FALSE, message=FALSE, warning=FALSE, Univariate_Plots4}

ggplot(redwine, aes(x = citric.acid)) +
geom_histogram(bindwith = .25, color = 'black') +
geom_vline(aes(xintercept = mean(citric.acid)), color = 'red', linetype = 'dashed',
           size = 1)


```
Citric Acid is positively correlated with the quality of red wines. We can see
here that the mean is around 0.25 g/L, and this is not a normal histogram at all,
but an entirely different shape.

It will be interesting to look at a multivariate analysis of this chart with 
regards to quality.


```{r echo=FALSE, message=FALSE, warning=FALSE, Univariate_Plots5}

ggplot(redwine, aes(x = volatile.acidity)) +
geom_histogram(bindwith = .25, color = 'black') +
geom_vline(aes(xintercept = mean(volatile.acidity)), color = 'red', linetype = 'dashed',
           size = 1)


```
Volatile Acidity is important, as we find out later, because it is negatively
associated with red wine quality. Here, we can see the mean is around 0.5 g/L.

This looks like a normal curve with a right tail. It will be interesting to look
at a multivariate analysis of this chart with regards to quality.

# Univariate Analysis

### What is the structure of your dataset?

This data set is structured with 11 numeric variables and 1 categorical
variable (quality). There is also the primary key in a column by itself. To be
more specific, all of the variables except for quality are numeric and real.
However, the variable for quality is numeric and an integer. I will have to
transform it, but it should work very well.

### What is/are the main feature(s) of interest in your dataset?

The categorical variable called quality could be useful when doing a
multivariate analysis. Density, pH, sulphites and alcohol content are tracked.
In addition, fixed acidity, volatile acidity, citric acid, residual sugar,
chlorides, free sulfur dioxide and total sulfur dioxide are tracked.

I am not sure what others would be interested in, but I am interested in how 
acids affect things like pH and density. I already know that more acid means
less pH, but I would like to see it in action with the data. I want to check and
see if more citric acid makes more fixed acidity.

Others might be interested in how alcohol content, sulfates, residual sugars, 
citric acid and even pH affect the quality rating. So we will look into that as
well.


### What other features in the dataset do you think will help support your investigation into your feature(s) of interest?

I really like how quality is an integer. This makes it a candidate for helping
out with multivariate analysis.

### Did you create any new variables from existing variables in the dataset?

I did not create any new variables.

### Of the features you investigated, were there any unusual distributions? Did you perform any operations on the data to tidy, adjust, or change the form of the data? If so, why did you do this?

Two of the variables, residual sugar and chlorides, have long tails on the right
and I would prefer to chop those off if I were to use those graphs in a
presentation.


# Bivariate Plots Section

### These are the correlation values of that data. From this, there are a few relationships we can look into.

```{r echo=FALSE, Bivariate_Plots}

cor(redwine)

```
Here, we can see that fixed acidity is negatively correlated with volatile
acidity and positively correlated with citric acid. Acidity in general
seems to be correlated with density, and also negatively correlated with pH.

What is most correlated with quality? Well, that would be alcohol content,
citric acid and sulfates. In addition, volatile acidity is associated with
poor quality. Residual sugars are interesting to glance at for a look at how
sweetness affects red wine quality.





### Bivarate Plots

```{r echo=FALSE, Bivariate_Plots2}

#redwine$fixed.acidity <- factor(redwine$fixed.acidity)
#redwine$citric.acid <- factor(redwine$citric.acid)

ggplot(redwine, aes(x =fixed.acidity, y = citric.acid)) +
         geom_point()

redwine$density = factor(redwine$density)

ggplot(redwine, aes(x = fixed.acidity, y = density)) +
         geom_point()

ggplot(redwine, aes(x = fixed.acidity, y = volatile.acidity)) +
         geom_point()

ggplot(redwine, aes(x = fixed.acidity, y = pH)) +
  geom_point()

ggplot(redwine, aes(x = alcohol, y = quality)) +
        geom_jitter(width=0, height=0.35)

ggplot(redwine, aes(x = residual.sugar, y = quality)) +
        geom_jitter(width=0, height=0.35)

ggplot(redwine, aes(x = citric.acid, y = quality)) +
        geom_jitter(width=0, height=0.35)

ggplot(redwine, aes(x = sulphates, y = quality)) +
        geom_jitter(width=0, height=0.35)

ggplot(redwine, aes(x = volatile.acidity, y = quality)) +
        geom_jitter(width=0, height=0.35)



```

### Descriptions of the scatterplots

It looks like citric acid is associated with fixed acidity, with a correlation
of .6717 . Fixed acidity is also positively correlated with density, with a 
correlation value of .6680 . 

What about citric acid vs. volatile acidity? We notice a negative correlation
here of -.5525, which gives us a clue to the nature of "volatile acidity" - it 
probably comes from something other than citric acid.

Keep reading for more information on the charts regarding quality.




# Bivariate Analysis

### Talk about some of the relationships you observed in this part of the investigation. How did the feature(s) of interest vary with other features in the dataset?

I showed a scatterplot of fixed acidity vs. citric acid to see if they were 
correlated, and they were, in a positive manner. Similarly, fixed acidity was 
positively correlated with density. 

These were scientific experiments to see if increasing citric acid increased
fixed acidity. I made the hypothesis that it would. In addition, I wanted to 
see if acids were denser than neutral substances, in general, and they were.

Interestingly enough, citric acid was negatively correlated with volatile
acidity.

Volatile acidity is negatively associated with fixed acidity, and I noticed that
pH tends to lower the more acids are added, which makes sense.

Wine sellers might be interested in the following:

Mean alcohol content seemed to increase as quality increased. Mean residual
sugar seemed to decrease quality. It is unclear to me how citric acid affects 
quality from the scatterplot, but I can tell that the mean sulfate content of 
high-quality wines is higher. Finally, higher volatile acidity was associated 
with lower-quality wines, which was surprising because my friend in the wine
industry said no one cared about volatile acidity content.

I will do all five of these again as a multivariate analysis.

### Did you observe any interesting relationships between the other features (not the main feature(s) of interest)?

Quality was most highly correlated with sulfates and alcohol content, and most
negatively associated with volatile acidity.


### What was the strongest relationship you found?

Fixed Acidity vs. Density was one of the strongest relationships I found, which
is quite interesting! Could there be something to this?

In addition, there were other high correlations, like sulfates vs. total
sulfur dioxide and fixed acidity vs. citric acid that were strong but expected.
There was also a strong relationship between pH and fixed acidity, which makes 
sense as the more acidic something is, the lower its pH, by law.

# Multivariate Plots Section

```{r echo=FALSE, Multivariate_Plots}

redwine2 <- redwine

redwine2$quality <- as.factor(redwine$quality)

ggpairs(data=redwine2, columns = 11:12, mapping = aes(color=quality))

```

### Description of Sulphates vs. Alcohol Content with Regards to Quality

This chart shows a scatterplot of sulfates vs. alcohol content with the
additional layer of quality added on top. From the alcohol chart in particular,
we notice that the highest quality wines have a higher alcohol content. For
sulfates, we notice the same trend - more sulfates means a higher-quality red 
wine.

As far as the correlation goes, for either terrible (3) or amazing (8) wines,
this is where sulfates were most negatively correlated with alcohol content, 
with terrible wines having a correlation as small as -.416

This last part seems like something very subtle, but is it usable information
if I can grasp my head around it?


```{r echo=FALSE, message=FALSE, warning=FALSE, Multivariate_Plots2}

alc_mean <- redwine %>%
  group_by(quality) %>%
  summarize(mean_val = mean(alcohol))

ggplot(redwine, aes(x = alcohol)) +
geom_histogram(bindwith = .25, color = 'black') +
geom_vline(aes(xintercept = mean(alcohol)), color = 'red', linetype = 'dashed',
           size = 1) +
  facet_grid(quality ~ .) +
geom_vline(data = alc_mean, aes(xintercept = mean_val, color=as.factor(quality)), linetype = 'dashed',
           size = 1) +
  facet_grid(quality ~ .)


```

### Description of Alcohol Content in Red Wines with Regards to Quality

Here, we can see that the higher-quality red wines have a higher alcohol
content. On this graph and the ones following, the mean of the entire batch
will be a red line going down the chart, with the mean lines of each quality
level also showing. The conclusion here would be to make your wine have as much
alcohol content as possible for a high rating. The mean alcohol rating of a 
high-quality red wine is around 12%.

```{r echo=FALSE, message=FALSE, warning=FALSE, Multivariate_Plots3}

ca_mean <- redwine %>%
  group_by(quality) %>%
  summarize(mean_val = mean(citric.acid))

ggplot(redwine, aes(x = citric.acid)) +
geom_histogram(bindwith = .25, color = 'black') +
geom_vline(aes(xintercept = mean(citric.acid)), color = 'red', linetype = 'dashed',
           size = 1) +
  facet_grid(quality ~ .) +
geom_vline(data = ca_mean, aes(xintercept = mean_val, color=as.factor(quality)), linetype = 'dashed',
           size = 1) +
  facet_grid(quality ~ .)


```

### Description of Citric Acid in Red Wines with Regards to Quality

We can see here that a higher mean of citric acid is associated with higher-
quality red wines. You can see that the mean citric acid of a high-quality wine
is around .4 g/L.


```{r echo=FALSE, message=FALSE, warning=FALSE, Multivariate_Plots4}

va_mean <- redwine %>%
  group_by(quality) %>%
  summarize(mean_val = mean(volatile.acidity))

ggplot(redwine, aes(x = volatile.acidity)) +
geom_histogram(bindwith = .25, color = 'black') +
geom_vline(aes(xintercept = mean(volatile.acidity)), color = 'red', linetype = 'dashed',
           size = 1) +
  facet_grid(quality ~ .) +
geom_vline(data = va_mean, aes(xintercept = mean_val, color=as.factor(quality)), linetype = 'dashed',
           size = 1) +
  facet_grid(quality ~ .)


```

### Description of Volatile Acidity in Red Wines with Regards to Quality

We can see here that volatile acidity is, in general, less in higher-quality
wines.

```{r echo=FALSE, message=FALSE, warning=FALSE, Multivariate_Plots5}

rs_mean <- redwine %>%
  group_by(quality) %>%
  summarize(mean_val = mean(residual.sugar))

ggplot(redwine, aes(x = residual.sugar)) +
geom_histogram(bindwith = .25, color = 'black') +
geom_vline(aes(xintercept = mean(residual.sugar)), color = 'red', linetype = 'dashed',
           size = 1) +
  facet_grid(quality ~ .) +
geom_vline(data = rs_mean, aes(xintercept = mean_val, color=as.factor(quality)), linetype = 'dashed',
           size = 1) +
  facet_grid(quality ~ .) +
  xlim(1.5,3)


```

### Description of Residual Sugar in Red Wines with Regards to Quality

We can see here that a higher mean of residual sugar is associated with the
extremities of wine quality. Sugar content jumps at 7, but goes back down to
slightly above the mean at quality level 8.

```{r echo=FALSE, message=FALSE, warning=FALSE, Multivariate_Plots6}

sul_mean <- redwine %>%
  group_by(quality) %>%
  summarize(mean_val = mean(sulphates))

ggplot(redwine, aes(x = sulphates)) +
geom_histogram(bindwith = .25, color = 'black') +
geom_vline(aes(xintercept = mean(sulphates)), color = 'red', linetype = 'dashed',
           size = 1) +
  facet_grid(quality ~ .) +
geom_vline(data = sul_mean, aes(xintercept = mean_val, color=as.factor(quality)), linetype = 'dashed',
           size = 1) +
  facet_grid(quality ~ .)


```
### Description of Sulfates in Red Wines with Regards to Quality

Here, we can see that higher-quality wines tend to have a higher sulfate count.
In fact, the mean of quality 8 wines is around .75 mg/L.


```{r echo=FALSE, message=FALSE, warning=FALSE, Multivariate_Plots7}

vola_mean <- redwine %>%
  group_by(quality) %>%
  summarize(mean_val = mean(volatile.acidity))

ggplot(redwine, aes(x = volatile.acidity)) +
geom_histogram(bindwith = .25, color = 'black') +
geom_vline(aes(xintercept = mean(volatile.acidity)), color = 'red', linetype = 'dashed',
           size = 1) +
  facet_grid(quality ~ .) +
geom_vline(data = vola_mean, aes(xintercept = mean_val, color=as.factor(quality)), linetype = 'dashed',
           size = 1) +
  facet_grid(quality ~ .)


```
### Volatile ACidity for Different Qualities of Wine

This chart shows how the mean of volatile acidity for different tiers of wine
quality goes down the higher one goes, until there is a slight uptick right at
level 8.


# Multivariate Analysis

### Talk about some of the relationships you observed in this part of the investigation. Were there features that strengthened each other in terms of looking at your feature(s) of interest?

I wanted to try looking at sulfates and alcohol content in relation to quality
one more time with a multivariate analysis. These two things were the most
correlated with quality, after all.

In the graphs produced you can look at the alcohol and phosphate distributions 
for different quality ratings.

The scatterplot featuring sulfates vs. alcohol content is there as well, with
quality layered on as different colors.

### Were there any interesting or surprising interactions between features?

I was surprised that mid-quality (5) wines were so closely packed around their
mean sulfate content and alcohol content. I am also surprised that alcohol and
sulfates are hardly correlated at all! This means that they are separate, 
independent features that could predict quality of wines in an algorithm.


------

# Final Plots and Summary

### Residual Sugar in Red Wines

```{r echo=FALSE, warning=FALSE, message=FALSE, Plot_One2}

qplot(x = residual.sugar, data = redwine2,
            xlab = 'Residual Sugar in Red Wines (g/L',
            ylab = 'Number of Red Wines',
            color = I('black'), fill = I('#990099'),
            title = "Residual Sugars in Red Wine",
            xlim = c(0, 8.5))

```

### Description

This plot shows the residual sugars in red wines. IF you wanted to fit in,
you might make your wine have around 2.539 grams per liter of residual sugars, 
which is the mean. This particular chart had a long right tail, which I chopped
off in order to make the graph more centralized.

### Fixed Acidity vs. Density in Red Wines

```{r echo=FALSE, message=FALSE, warning=FALSE, Plot_Two2}

ggplot(redwine, aes(x= fixed.acidity, y = density)) +
         geom_point() +
ggtitle("Fixed Acidity vs. Density in Red Wines")


```

### Description

This graph shows fixed acidity vs. density. I was interested in seeing if there
was a relationship, and there was a positive correlation of .668 . I checked out
another online data set and found out that this was the case there as well.


### Volatile ACidity vs. Quality Rating

```{r echo=FALSE, message=FALSE, warning=FALSE, Plot_Three3}

vola_mean <- redwine %>%
  group_by(quality) %>%
  summarize(mean_val = mean(volatile.acidity))

ggplot(redwine, aes(x = volatile.acidity)) +
geom_histogram(bindwith = .25, color = 'black') +
geom_vline(aes(xintercept = mean(volatile.acidity)), color = 'red', linetype = 'dashed',
           size = 1) +
  facet_grid(quality ~ .) +
geom_vline(data = vola_mean, aes(xintercept = mean_val, color=as.factor(quality)), linetype = 'dashed',
           size = 1) +
  facet_grid(quality ~ .) +
labs(title = "Volatile Acidity vs. Quality Rating in Red Wines") +
xlab("Volatile Acidity (grams per liter)")


```

### Description

This graph looks at the relationships between volatile acidity and quality
rating. Although the wine-industry insider I knew thought that volatile acidity
did not matter for wine quality, it was the most negatively associated with
wine quality out of all of the variables explored.

------

# Reflection

I found out that a multivariate analysis was better for exploring the
relationship between things such as sulfates vs. quality than a scatterplot.
I could not get a feel for those scatterplots that I made earlier in the
exploration. I eventually learned how to use jitter to clear this up a bit.

I was personally interested in using this data set to learn a bit about
chemistry and physics, while I think others might be interested in using it
to learn what kinds of wines have a high quality rating. So I did a bit of both
in the exploration.

The first two scatterplots I made, citric acid vs. fixed acidity and fixed
acidity vs. density went well. In addition, the mass-production of 
correlation values as well as univariate charts went very well. Those gave
a decent overview of the data, along with the data summary. I thought the
multivariate analysis went well, but it is my first time doing one, and I am
not sure.

I was surprised at the relationship between acidity and density. In the future,
I would like to do more multivariate analysis with the quality variable, and see
what I can find.